---
title: "Kern_fallowing_time_series"
author: "James Cunningham"
date: "2024-10-11"
output: html_document
---

# Introduction

This script generates a series of raster .tif files that summarize agricultural activity in Kern County, CA over time (1985–2020) using land cover data sourced from FMMP and USGS. 

Output Rasters:
Last year cropped: raster_max_Kern_1985_2020.tif
Earliest year cropped: raster_min_Kern_1985_2020.tif
Agriculture occurrence count: ag_count_Kern_1985_2020.tif
Transitions away from agriculture: transition_count_1985_2020.tif

Stored Paths: These raster files are saved in the processed folder for use in the second script Kern_fallowing_extract.Rmd

Second Script (Kern_fallowing_extract.Rmd): Reads raster files and extracts values (using a mode calculation) to populate the fields of the 1990 CDWR shapefile.

```{r setup, include=FALSE}

# Automatically install required packages if not already installed
required_packages <- c("terra", "sf", "dplyr", "here")
for(pkg in required_packages) {
  if(!require(pkg, character.only = TRUE)) {
    install.packages(pkg, dependencies = TRUE)
    library(pkg, character.only = TRUE)
  }
}

# Load needed libraries
library(terra)
library(dplyr)
library(sf)
library(here)

# Initialize an empty vector to store raster paths
raster_paths <- c()
```


# Part 1: Rasterize the GIRAS (1970-1985) land cover data

```{r}

# Read in shapefiles
g35120 <- st_read(here("data", "raw", "GIRAS", "g35120.shp"))
g35118 <- st_read(here("data", "raw", "GIRAS", "g35118.shp"))
g35116 <- st_read(here("data", "raw", "GIRAS", "g35116.shp"))
g34118 <- st_read(here("data", "raw", "GIRAS", "g34118.shp"))
g34116 <- st_read(here("data", "raw", "GIRAS", "g34116.shp"))
g33120 <- st_read(here("data", "raw", "GIRAS", "g33120.shp"))
g33118 <- st_read(here("data", "raw", "GIRAS", "g33118.shp"))
g33116 <- st_read(here("data", "raw", "GIRAS", "g33116.shp"))

# Combine shapefiles
combined_shape <- rbind(g35120, g35118, g35116, g34118, g34116, g33120, g33118, g33116)

# Write it to disk
st_write(combined_shape, here("data", "intermediate", "combined_shapefile.shp"), append=FALSE)

# Transform to EPSG 3309 (to match FMMP default CRS)
combined_shape_transformed <- st_transform(combined_shape, 3309)

# Read in FMMP Kern County file to clip, allows FMMP rasters to overlap
kern1988_clip <- st_read(here("data", "raw", "FMMP", "kern1988.shp"))

# Perform the intersection, which finds the overlapping area between GIRAS and FMMP
clipped_shape <- st_intersection(combined_shape_transformed, kern1988_clip)

# Filter to retain only attributes from GIRAS file, as second part of clip
clipped_shape <- clipped_shape[, names(combined_shape_transformed)]

# Write out the clipped GIRAS shapefile
st_write(clipped_shape, here("data", "intermediate", "giras_kern1985.shp"), append=FALSE)

# Transform the shapefile to EPSG:26941
kern_1985_transformed <- st_transform(clipped_shape, 26941)

# Reclassify polygons, wherever crops exist assign that polygon the year value, GIRAS is 1985, see Readme for more info
kern_1985_reclassified <- kern_1985_transformed %>%
  mutate(year_cropped = case_when(
    LUCODE %in% c(21, 22, 24) ~ 1985,
    LUCODE %in% c(11, 12, 13, 14, 15, 16, 17, 21, 22, 23, 24, 31, 32, 33, 41, 42, 
                  43, 51, 52, 53, 61, 62, 71, 73, 74, 75, 76, 77) ~ NA_integer_,
    TRUE ~ NA_integer_
  ))

# Define extent, values are from FMMP Kern shapefiles
xmin = 2163844.24083589
xmax = 2401524.24083589
ymin = -593.637201080564
ymax = 116706.8484412

ext <- ext(xmin, xmax, ymin, ymax)

# Rasterize, since units are meters in the Projected CRS EPSG: 26941, res=10 means 10m resolution
rast_template <- rast(ext, res=10)
rasterized_1985 <- rasterize(kern_1985_reclassified, rast_template, field="year_cropped")

# Manually set CRS of raster output
crs(rasterized_1985) <- "EPSG:26941"

# Write out the 1985 raster
output_path <- here("data", "intermediate", "rasterized_1985.tif")
writeRaster(rasterized_1985, output_path, overwrite=TRUE)

# Store the raster output path
raster_paths <- c(raster_paths, output_path)

```


# Part 2: Rasterize 1988-2020 FMMP shapefiles

```{r}


# Listing the years of Kern data
years <- c(1988, 1990, 1992, 1994, 1996, 1998, 2000, 2002, 2004, 2006, 2008, 2010, 2012, 2014, 2016, 2018, 2020)

# Loop over each year
for(year in years) {
  
  # Construct shapefile path
  shapefile_path <- here("data", "raw", "FMMP", paste0("kern", year, ".shp"))
  
  # Read the shapefile
  polygon_sf <- st_read(shapefile_path)
  
  # Check the year to assign and then transform the CRS
  if(year <= 2012) {
    st_crs(polygon_sf) <- 3309  # Set the CRS for years 1988-2012, note: FMMP metadata indicates 3309 for these years
  } else {
    st_crs(polygon_sf) <- 3310  # Set the CRS for years 2014-2020, note: FMMP metadata indicates 3309 for these years
  }
  
  # Transform the shapefile to EPSG:26941
  polygon_sf_transformed <- st_transform(polygon_sf, 26941)
  
  # Reclassify polygons, wherever crops exist assign that polygon the year value
  polygon_sf_reclassified <- polygon_sf_transformed %>%
    mutate(year_cropped = case_when(
      polygon_ty %in% c('P', 'U', 'L', 'S', 'I', 'N') ~ year,
      polygon_ty %in% c('Cl', 'D', 'G', 'nv', 'R', 'sAC', 'V', 'W', 'X') ~ NA_integer_,
      TRUE ~ NA_integer_
    ))
  
  # Define extent, this is to be inclusive of the range of extents for each shapefile
  xmin = 2163844.24083589
  xmax = 2401524.24083589
  ymin = -593.637201080564
  ymax = 116706.8484412
  
  ext <- ext(xmin, xmax, ymin, ymax)
  
  # Rasterize, since units are meters in the Projected CRS EPSG: 26941, res=10 means 10m resolution
  rast_template <- rast(ext, res=10)
  rasterized <- rasterize(polygon_sf_reclassified, rast_template, field="year_cropped")
  
  # Manually set CRS of raster output
  crs(rasterized) <- "EPSG:26941"
  
  # Write out the raster using here()
  output_path <- here("data", "intermediate", paste0("rasterized_", year, ".tif"))
  writeRaster(rasterized, output_path, overwrite=TRUE)
  
  # Store the raster output path
  raster_paths <- c(raster_paths, output_path)
  
}

```


# Part 3: Raster Algebra setup

```{r}


# Read all rasters into a multi-layer SpatRaster
raster_stack <- rast(raster_paths)

# Check if all rasters are loaded correctly, should be 18, 17 FMMP, 1 GIRAS
print(nlyr(raster_stack))

# Save as R object to use in other scripts if needed
saveRDS(raster_stack, file = here("data", "intermediate", "raster_stack.rds"))

```

# Raster Calculations

```{r}
### Max raster ###

# Calculate max for each cell across the stack
max_raster <- app(raster_stack, fun = max, na.rm = TRUE)

# Check unique values, making sure you see an assortment of years
print(unique(values(max_raster)))

# Write the max raster to disk using here()
writeRaster(max_raster, here("data", "processed", "raster_max_Kern_1985_2020.tif"), overwrite=TRUE)

```


```{r}
### Min raster ###

# Calculate min for each cell across the stack
min_raster <- app(raster_stack, fun = min, na.rm = TRUE)

# Check unique values, making sure you see an assortment of years
print(unique(values(min_raster)))

# Write the min raster to disk using here()
writeRaster(min_raster, here("data", "processed", "raster_min_Kern_1985_2020.tif"), overwrite=TRUE)

```

```{r}
### How many times has each cell had ag?

# Function to count non-NA values (i.e., agriculture occurrences)
count_agriculture <- function(x) {
  sum(!is.na(x))
}

# Apply this function across the stack
agriculture_count_raster <- app(raster_stack, count_agriculture)

# Replace cells with 0 count with NA
agriculture_count_raster[agriculture_count_raster == 0] <- NA

# Set the CRS of the output raster
crs(agriculture_count_raster) <- "EPSG:26941"

# Save the output raster
writeRaster(agriculture_count_raster, here("data", "processed", "ag_count_Kern_1985_2020.tif"), overwrite=TRUE)


```


```{r}
### How many times has the cell transitioned away from ag? ###

# Note: The years in raster_stack need to go from oldest to newest for this to work.

# Function to count transitions from agriculture to non-agriculture across raster stack, so comparing individual cells across time.
count_transitions <- function(cell_values) {
  # Initialize counter for transitions
  transitions <- 0
  
  # Loop through each cell value to compare with the next value (ex.1988 with 1990)
  for (i in 1:(length(cell_values) - 1)) {
    # Check if the current cell value is not NA (indicates it is year indicating agriculture) 
    # followed by NA (indicating no agriculture), this is counted as a transition away from ag
    # If true, the transitions counter increased by 1
    if (!is.na(cell_values[i]) && is.na(cell_values[i + 1])) {
      transitions <- transitions + 1
    }
  }
  # Cell values are the total number of tranistions for that cell
  return(transitions)
}

# Apply the function across the stack
transition_count_raster <- app(raster_stack, count_transitions)

# Set the CRS of the output raster
crs(transition_count_raster) <- "EPSG:26941"

# Save the output raster
writeRaster(transition_count_raster, here("data", "processed", "transition_count_Kern_1985_2020.tif"), overwrite=TRUE)

```


