---
title: "Kern_fallowing_extract"
author: "James Cunningham"
date: "2024-10-11"
output: html_document
---

# Introduction

This second script reads raster files (first year cropped, last year cropped, transition count, agriculture count) and extracts values (using a mode calculation) to populate the fields of the 1990 CDWR shapefile for Kern County, CA.

The extraction matches the CRS from the .tif files to CDWR, then replaces NA values with -9999 and does a mode calculation such that a field is populated by the most present pixel value under it. 


# Setup

```{r setup, include=FALSE}

# Automatically install required packages if not already installed
required_packages <- c("terra", "sf", "dplyr", "here")
for(pkg in required_packages) {
  if(!require(pkg, character.only = TRUE)) {
    install.packages(pkg, dependencies = TRUE)
    library(pkg, character.only = TRUE)
  }
}

# Load necessary libraries
library(terra)
library(sf)
library(dplyr)
library(here)
```

# Extraction

```{r}

# Define file paths using here()
raster_max_path <- here("data", "processed", "raster_max_Kern_1985_2020.tif")
transition_count_path <- here("data", "processed", "transition_count_Kern_1985_2020.tif")
ag_count_path <- here("data", "processed", "ag_count_Kern_1985_2020.tif")
min_year_path <- here("data", "processed", "raster_min_Kern_1985_2020.tif")

shapefile_path <- here("data", "raw", "i15_LandUse_Kern1990", "i15_LandUse_Kern1990.shp")

# Load the rasters
raster_max <- rast(raster_max_path)
transition_count_raster <- rast(transition_count_path)
ag_count_raster <- rast(ag_count_path)
min_year_raster <- rast(min_year_path)

# Load the shapefile
cdwr_fields <- st_read(shapefile_path)

# Ensure all rasters match the CRS and extent of the CDWR shapefile
raster_crs <- st_crs(cdwr_fields)

if (crs(raster_max) != raster_crs) {
  raster_max <- project(raster_max, raster_crs$proj4string)
}

if (crs(transition_count_raster) != raster_crs) {
  transition_count_raster <- project(transition_count_raster, raster_crs$proj4string)
}

if (crs(ag_count_raster) != raster_crs) {
  ag_count_raster <- project(ag_count_raster, raster_crs$proj4string)
}

if (crs(min_year_raster) != raster_crs) {
  min_year_raster <- project(min_year_raster, raster_crs$proj4string)
}

# Replace NA values with -9999 in each raster
# This allows the mode calculation to return a value of -9999 where non-majority agriculture
# land cover is not recorded under a CDWR polygon
# This helps distinguish areas with no data from those with valid agricultural records under a CDWR polygon
# Note: ifel is like ifelse, but the Terra version optimized for rasters
raster_max <- ifel(is.na(raster_max), -9999, raster_max)
transition_count_raster <- ifel(is.na(transition_count_raster), -9999, transition_count_raster)
ag_count_raster <- ifel(is.na(ag_count_raster), -9999, ag_count_raster)
min_year_raster <- ifel(is.na(min_year_raster), -9999, min_year_raster)

# Define a function to calculate the mode, handling NA values
# This function will be used in the extract() function to to compute the most common raster value
# for each CDWR polygon.
calculate_mode <- function(x) {
  x <- na.omit(x)  # Remove NA values
  ux <- unique(x)  # Get the unique raster cell values in the area of a polygon (vector x)
  if (length(ux) == 0) {
    return(NA)  # Return NA if all values are NA
  } else {
    # Count the occurrences of each unique value in x, and return the most frequent one
    mode_value <- ux[which.max(tabulate(match(x, ux)))]
    return(as.integer(mode_value))  # Return the mode as an integer
  }
}

# Extract raster values and calculate mode for each polygon in the CDWR shapefile
# Applies the 'calculate_mode()' function to summarize the most frequent value for each polygon
# The resulting mode values are stored in new fields in the 'cdwr_fields' shapefile
cdwr_fields$falevt <- extract(transition_count_raster, vect(cdwr_fields), fun = calculate_mode)[, 2]
cdwr_fields$totyc <- extract(ag_count_raster, vect(cdwr_fields), fun = calculate_mode)[, 2]
cdwr_fields$frstyc <- extract(min_year_raster, vect(cdwr_fields), fun = calculate_mode)[, 2]
cdwr_fields$lastyc <- extract(raster_max, vect(cdwr_fields), fun = calculate_mode)[, 2]

# Save the updated shapefile
output_path <- here("data", "processed", "Kern1990_CDWR_with_FMMP_GIRAS.shp")
st_write(cdwr_fields, output_path, delete_layer = TRUE)

```


